---
title: "perturbR cookbook"
author: "Simon Dirmeier"
date: "`r Sys.Date()`"
output: BiocStyle::html_document
bibliography: knockdown.bib
vignette: >
  %\VignetteIndexEntry{Knockdown cookbook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r style, echo=FALSE, results='asis'}
  BiocStyle::markdown()
```

```{r setup, include=FALSE}
  knitr::opts_chunk$set(echo = TRUE)
  options(warn = -1)
  library(data.table)
  library(dplyr)
  library(dtplyr)
  library(methods)
  library(perturbR)
  data(rnaiscreen)
  my.data <- rnaiscreen@.data %>%
    dplyr::filter(Condition=="V1") %>%
    dplyr::select(Condition, Replicate, Plate, RowIdx, ColIdx,
                  GeneSymbol, Readout, Control, Perturbation)
```

# Introduction

`perturbR` does stage-wise analysis of large-scale genetic perturbation screens for integrated datasets consisting of several screens. The package provides various tools for normalisation, plotting and analysis. For single perturbation screens classical analyses using hypothesis testing are implemented. For multiple integrated perturbation screens we developed a hierarchical model that considers the variance between the different biological settings the data are derived from. Our model first estimates an overall relative size for all genes (hits). The resulting hit lists is then be further extended using a network propagation algorithm to correct for false negatives and positives.

# Tutorial

This tutorial walks you to the basic functionality of `perturbR` using a pan-pathogenic data set of several RNAi screening studies.

## Creating a raw data-set

You supposedly start with something like a `data.frame`. In order to create a
`knockdown-data`-set you need to transfrom it into a `data.table` with at least
the following columns:

```{r, include=TRUE, size="tiny"}
  print(my.data)
```

You can then use your data-set `my.data` and cast it to a `perturbation.data` object:

```{r, include=TRUE, eval=TRUE}
  my.perturbation.data <- methods::as(my.data, "perturbation.data")
```

Coercing your `data.table` to `perturbation.data` will automatically warn you if
your table is formatted wrongly.

## Working with `perturbation.data` S4 objects

For objects that inherit the abstract `perturbation.data` class, we provide basic
`select`, `filter`, `group_by` and `rbind` functionality:

```{r, eval=TRUE, include=TRUE}
  select(my.perturbation.data, GeneSymbol, Readout)
  rbind(filter(my.perturbation.data, Readout > 0),
        filter(my.perturbation.data, Readout < 0))
```

## Preprocessing

Before starting to work with the data we might just want to have a plain look how many genes and replicates we have.
We use one of the provided artificial data-sets for this.

```{r, eval=TRUE, include=TRUE, warning=FALSE, fig.align="center", fig.width=3, fig.height=4}
  data(rnaiscreen)
  plot(rnaiscreen)
```

### Normalisation

The `preprocess` method offers a collection of common techniques to normalize
RNAi-screening data. We just call it with the standard parameters, which takes the natural
logarithm on every plate and afterwards computes a more robust version of the standard Z-score.

```{r, eval=TRUE, include=TRUE, warning=FALSE, message=FALSE}
  v1.data.norm <- perturbR::preprocess(v1.data, normalize=c("log", "robust-z.score"))
```

The result shouldn't be too different from before since we use artificial data. But for completeness' sake let's do the full analysis. We do the same as
before (this time w/o plots, though)

```{r, eval=FALSE, include=TRUE}
  v1.data.norm.quality <- quality(v1.data.norm)
  pl <- plot(v1.data.norm.quality)

  plates <- plates(v1.data.norm)
  plot(plates[1], show.gene.names=T, gene.text.size=2)

  repls <- replicates(v1.data.norm)
  plot(repls[1], repls[2])
```

The result object has class `perturbation.normalized.data` which can be used for analysis.

In the example above we just computed the log and robust Z-scores for every
plate separately. Often, when we have for instance information of cell-counts,
cell-viability or expect spatial effects due to several batch effects other
normalization methods are more suitable. Furthermore when your data-set
comprises multiple different screens, you might want to normalize every screen
separately.

In this case you would first filter the data-set by single experiments. For our provided
filtering by `Condition` and `Screen` results int a single experiment;
*for your data this might be different though*.

```{r, eval=TRUE, include=T}
  v1.dat <- filter(rnaiscreen, Condition=="V1")
  v2.dat <- filter(rnaiscreen, Condition=="V2")
```

## Data-analysis

After preprocessing of raw RNAi data, analysis can either be done on the single
viral level using hypothesis tests or using a maximum
likelihood approach.

### Hypothesis testing

Suppose we normalized our data-set and centered the data around zero. One way to find significant genes
is to use a standard `t-test`:

```{r, eval=TRUE, include=TRUE, warning=FALSE, message=FALSE, fig.align="center", fig.width=4}
  v1.dat <- perturbation::filter(rnaiscreen, Virus=="V1")
  v1.res <- tstatistic(v1.dat)
  plot(v1.res)
```

Alternatively you can also use an iterative hyper-geometric test with `hyper.statistic` [@konig2007probability]. We also refer the interested reader to a recent model for hit prioritization using a penalized linear model by [@schmich2015gesper].

### Joint maximum likelihood inference of integrated data sets

If we want to analyse several pathogens add once and detect host factors common to all of those other appraoches might make more sense.
For this we developed a hierchical model that separately treats biological effects from different settings called `hm`.
We describe the phenotype/readout by a linear combination of fixed effects and random effects (inspired by [@ramo2014simultaneous]) as:

$$y_{vgst} = x_v \beta + \gamma_g  + \delta_{vg} + \zeta_t + \xi_{vt} + \epsilon_{vgst},$$
where $y_{vgst}$ is the phenotype for a virus $v$, gene $g$, siRNA $s$ and infection type $t$.
Since RNAi screens are prone to having large variance, we try to capture as much of the
variance using random effects for genes, viruses and the infection types. You call the
random effects model using:

```{r, eval=TRUE, include=TRUE, warning=FALSE, message=FALSE}
  rnaiscreen.normalized <- preprocess(rnaiscreen, normalize="robust-z.score")
  res                   <- hm(rnaiscreen.normalized, effect.size=0.01)
```

If we `plot` our result we are getting a list three different plots. The most informative one that shows the top 25 pen-pathogenic host factors:

```{r eval=TRUE, include=TRUE, fig.align="center", fig.width=4, warning=FALSE, message=FALSE}
  pl <- plot(res)
  print(pl[[1]])
```

Since our screen had only some few genes, we might want to *smooth* the effect from the random effects model using network diffusion. For that
we need a graph file and call the `diffuse` function. If we plot the results we get a nice visualisation of the graph and newly detected hits using
the diffusion. In our example we do not compute the diffusion, because loading the graph takes a lot memory, but go ahead a try yourself:


```{r eval=FALSE, include=TRUE, fig.align="center", fig.width=4, message=FALSE}
  graph.file <- system.file("extdata", "graph_file.tsv", package = "perturbR")
  diffu <- diffuse(res, method="mrw", path=graph.file, r=0.1)
  plot(diffu)
```
## Session info

```{r eval=TRUE, include=TRUE}
  sessionInfo()
```

## References
